---
title: "Final Project"
author: "Lane Riggs"
date: "2023-04-15"
output: "html_document"
---

_This is the final project for the HIST 8510 course. In this final project, I will be looking at data about migrant deaths at the border. U.S. Customs and Border Protection agents have significantly underrecorded immigrant deaths, with numbers likely twice as high as previously reported._ 

_The Government Accountability Office (GAO) stated "CBP has not collected and recorded, or reported to Congress, complete data on migrant deaths or disclosed limitations with the data it has reported."_

_Since 1998, the Border Patrol has recorded approximately 8,600 immigrant deaths at the border. If this data is correct, the actual number of immigrant deaths might have been twice as high since 1998._ 

```{r}

#first we are going to load the dataset and then rename it 

tucson.numbers <- data.9B69e

#I found this dataset from an article posted to Forbes in 2022. 

```

```{r}

#we will first graph the underrepresented deaths at the border

#colors are chosen to represent the Border Patrol colors

tucson.numbers.df <- data.frame(reported = c(63, 84, 73, 58, 61), not.reported = c(86, 66, 71, 65, 72))

barplot(tucson.numbers.df$not.reported,
        tucson.numbers.df$reported,
        main = "Actual Deaths at the Border",
        xlab = "Deaths",
        col = c("green", "blue")
        )

legend("topleft",
       c("Reported", "Not Reported"),
       fill = c("green", "blue")
       )

```
>



```{r}



````


